Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-09-25 manuscripts/reply/reply-abstract.tex
Incorporating some edits to proofs.

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that a common process affected
the populations.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model choice method (\msb) to estimate such a pattern of simultaneous
divergence across taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa when prior uncertainty about divergence times
is represented by a uniform distribution.
However, the papers differ about the best explanation and solution for this
problem.
\citet{Oaks2012} suggested the method's behavior was caused by the strong
weight of uniformly distributed priors on divergence times leading to smaller
marginal likelihoods (and thus smaller posterior probabilities) of models with
more divergence time parameters (Hypothesis~\ref{hypML}); they proposed
alternative prior probability distributions to avoid such strongly weighted
posteriors.
\citet{Hickerson2013} suggested numerical approximation error causes \msb
analyses to be biased toward models of clustered divergences because the
method's rejection algorithm is unable to adequately sample the parameter space
of richer models within reasonable computational limits when using broad
uniform priors on divergence times (Hypothesis~\ref{hypError}).
As a potential solution, they proposed a model-averaging approach that uses
narrow, empirically informed uniform priors.
Here, we use analyses of simulated and empirical data to demonstrate that the
approach of \citet{Hickerson2013} does not mitigate the method's tendency to
erroneously support models of highly clustered divergences, and is dangerous in
the sense that the empirically-derived uniform priors often exclude from
consideration the true values of the divergence-time parameters.
Our results also show that the tendency of \msb analyses to support models of
shared divergences is primarily due to Hypothesis~\ref{hypML}, whereas
Hypothesis~\ref{hypError} is an untenable explanation for the bias.
Overall, this series of papers demonstrate that if our prior assumptions place
too much weight in unlikely regions of parameter space such that the exact
posterior supports the wrong model of evolutionary history, no amount of
computation can rescue our inference. 
Fortunately, as predicted by fundamental principles of Bayesian model choice,
more flexible distributions that accommodate prior uncertainty about parameters
without placing excessive weight in vast regions of parameter space with low
likelihood increase the method's robustness and power to detect temporal
variation in divergences.

>>>>

Establishing that a set of population-splitting events occurred at the same
time can be a potentially persuasive argument that a common process affected
the populations.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model-choice method (\msb) to estimate such a pattern of simultaneous
divergence across taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa when prior uncertainty about divergence times
is represented by a uniform distribution.
However, the papers differ about the best explanation and solution for this
problem.
\citet{Oaks2012} suggested the method's behavior was caused by the strong
weight of uniformly distributed priors on divergence times leading to smaller
marginal likelihoods (and thus smaller posterior probabilities) of models with
more divergence-time parameters (Hypothesis~\ref{hypML}); they proposed
alternative prior probability distributions to avoid such strongly weighted
posteriors.
\citet{Hickerson2013} suggested numerical approximation error causes \msb
analyses to be biased toward models of clustered divergences because the
method's rejection algorithm is unable to adequately sample the parameter space
of richer models within reasonable computational limits when using broad
uniform priors on divergence times (Hypothesis~\ref{hypError}).
As a potential solution, they proposed a model-averaging approach that uses
narrow, empirically informed uniform priors.
Here, we use analyses of simulated and empirical data to demonstrate that the
approach of \citet{Hickerson2013} does not mitigate the method's tendency to
erroneously support models of highly clustered divergences, and is dangerous in
the sense that the empirically-derived uniform priors often exclude from
consideration the true values of the divergence-time parameters.
Our results also show that the tendency of \msb analyses to support models of
shared divergences is primarily due to Hypothesis~\ref{hypML}, whereas
Hypothesis~\ref{hypError} is an untenable explanation for the bias.
Overall, this series of papers demonstrate that if our prior assumptions place
too much weight in unlikely regions of parameter space such that the exact
posterior supports the wrong model of evolutionary history, no amount of
computation can rescue our inference. 
Fortunately, as predicted by fundamental principles of Bayesian model choice,
more flexible distributions that accommodate prior uncertainty about parameters
without placing excessive weight in vast regions of parameter space with low
likelihood increase the method's robustness and power to detect temporal
variation in divergences.

### 2014-09-25 manuscripts/reply/reply-main.tex
Incorporating some edits to proofs.

\section{Introduction}
Biogeographers frequently seek to explain population and species
differentiation on geographical phenomena.
Establishing that a set of population splitting events occurred
at the same time can be a potentially persuasive argument that a set of taxa
were affected by the same geographic events.
The approximate-Bayesian method, \msb, allows biogeographers to estimate the
probabilities of models in which multiple sets of taxa diverge at the same
time \citep{Hickerson2006,Huang2011}.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item Representing prior uncertainty about divergence-time parameters with a
       uniform distribution can lead to spurious support for models with few
       divergence events shared across taxa. Thus, the primary inference
       enabled by the approach is very sensitive to the priors on divergence
       times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{enumerate}[label=\textsl{\textbf{Hypothesis \arabic*)}},ref=\arabic*,align=left]
    \item \textsl{\textbf{Strongly weighted marginal likelihoods}}
        \citep{Oaks2012}:
        The uniform priors on divergence times lead to very small marginal
        likelihoods (and thus smaller posterior probabilities) of models with
        many divergence-time parameters.  The likelihood of these models is
        ``averaged'' over a much greater parameter space in which there is a
        large amount of prior weight and small probability of producing the
        data.  \citep{Jeffreys1939,Lindley1957}. \label{hypML}
    \item \textsl{\textbf{Numerical approximation error}} \citep{Hickerson2013}:
        Under broad uniform priors, the rejection algorithm implemented in \msb
        is unable to adequately sample the space of the models within
        reasonable computational time, which leads to bias toward models with
        fewer divergence-time parameters because they are better sampled.
        \label{hypError}
\end{enumerate}
In Hypothesis~\ref{hypError}, the problem is numerical approximation error due
to insufficient computation.
In this scenario, given data from taxa that diverged randomly through time, the
exact (true) posterior supports a model with many divergence-time parameters,
but we are unable to accurately approximate this posterior.
In Hypothesis~\ref{hypML}, the problem is more fundamental; given data from
taxa that diverged randomly through time, the exact posterior supports a model
with simultaneous divergences across taxa.
I.e., when accommodating prior uncertainty about divergence times with a
uniform distribution, the exact posterior from Bayes' rule leads us to the
wrong conclusion about evolutionary history.
Such posterior support for simultaneous divergence, even if ``correct'' from
the perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

@@

\begin{linenomath}
Whereas empirical Bayesian approaches can provide powerful methods for
parameter estimation, a theoretical justification for empirical Bayesian
approaches to model choice is questionable.
In Bayesian model choice, the primary goal is not to estimate parameters, but
to estimate the probabilities of candidate models.
In a simple example with two candidate models, $M_1$ and $M_2$, we can use
Bayes' rule to calculate the posterior probability of $M_1$ as
\begin{equation}
    p(M_1 \given \alignment{i}{}) = \frac{p(\alignment{i}{} \given
    M_1)p(M_1)}{p(\alignment{i}{} \given M_1)p(M_1) + p(\alignment{i}{} \given
    M_2)p(M_2)}.
    \label{eq:bayesmodelchoice}
\end{equation}
By comparing Equations \ref{eq:bayesrule} and \ref{eq:bayesmodelchoice}, we
see fundamental differences between Bayesian parameter estimation and
model choice.
\end{linenomath}

@@

However, \citet{Hickerson2013} conclude that the bias is caused by numerical
approximation error due to insufficient computation
(Hypothesis~\ref{hypError}).
They argue the widest of the three priors on divergence times used by
\citet{Oaks2012} would infrequently produce random samples of parameter values
with many independent population divergence times as recent as the estimated
gene divergence times presented in \citet{Oaks2012}.
% They also also suggest that Oaks et al.'s \citeyear{Oaks2012} estimated gene
% divergence times should be used to set the upper bound of the uniform prior on
% divergence times for the \msb analysis of the same sequence data.
% Given the gene-tree estimates were intended to provide only a rough comparison
% of the gene divergence times across the 22 taxa \citep{Oaks2012}, and our
% findings presented above regarding the behavior of \msb when empirically
% informed uniform priors are employed, such an approach should not be used.
However, this sampling-probability argument is based on some questionable
assumptions.
Oaks et al.'s (\citeyear{Oaks2012}) gene-tree estimates were intended to
provide only a rough comparison of the gene divergence times across the 22 taxa
and assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations per
generation for all taxa.
Furthermore, because the branch-length units of the gene trees are in millions
of years whereas the divergence-time prior of \msb is in generations,
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
More importantly, even if we assume (1) the arbitrary strict clock is correct,
(2) gene divergence times were estimated without error, and (3) all 22 taxa
have one-year generation times, Hickerson et al.'s (\citeyear{Hickerson2013})
argument actually demonstrates that the models used by \citet{Oaks2012} with
narrower priors on divergence times are densely populated with samples with
large numbers of divergence parameters with values younger than the estimated
gene divergence estimates.
Thus, if \citet{Hickerson2013} are correct, analyses under these narrow priors
should be much less biased toward clustered models of divergence.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrowest prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug in
\msb, and after \citet{Oaks2012} corrected the bug, there are many cases where
the narrowest prior performs slightly worse (see panels D--J of Figures 3 and
S12).

@@

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that often exclude the true
values of the divergence times.
Our results show that it is difficult to choose an uniformly distributed prior
on divergence times that is broad enough to confidently contain the true values
of parameters while being narrow enough to avoid strongly weighted and
misleading posterior support for models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection, because they can strongly influence the
posterior probabilities of the models we seek to compare.
No amount of computation can rescue our inference if our prior assumptions
place too much weight in unlikely regions of parameter space such that the
exact posterior supports the wrong model of evolutionary history.

>>>>

\section{Introduction}
Biogeographers frequently seek to explain population and species
differentiation on geographical phenomena.
Establishing that a set of population-splitting events occurred
at the same time can be a potentially persuasive argument that a set of taxa
were affected by the same geographic events.
The approximate-Bayesian method, \msb, allows biogeographers to estimate the
probabilities of models in which multiple sets of taxa diverge at the same
time \citep{Hickerson2006,Huang2011}.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item Representing prior uncertainty about divergence-time parameters with a
       uniform distribution can lead to spurious support for models with few
       divergence events shared across taxa. Thus, the primary inference
       enabled by the approach is very sensitive to the priors on divergence
       times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{enumerate}[label=\textsl{\textbf{Hypothesis \arabic*)}},ref=\arabic*,align=left]
    \item \textsl{\textbf{Strongly weighted marginal likelihoods}}
        \citep{Oaks2012}:
        The uniform priors on divergence times lead to very small marginal
        likelihoods (and thus smaller posterior probabilities) of models with
        many divergence-time parameters.  The likelihood of these models is
        ``averaged'' over a much greater parameter space in which there is a
        large amount of prior weight and small probability of producing the
        data.  \citep{Jeffreys1939,Lindley1957}. \label{hypML}
    \item \textsl{\textbf{Numerical-approximation error}} \citep{Hickerson2013}:
        Under broad uniform priors, the rejection algorithm implemented in \msb
        is unable to adequately sample the space of the models within
        reasonable computational time, which leads to bias toward models with
        fewer divergence-time parameters because they are better sampled.
        \label{hypError}
\end{enumerate}
In Hypothesis~\ref{hypError}, the problem is numerical-approximation error due
to insufficient computation.
In this scenario, given data from taxa that diverged randomly through time, the
exact (true) posterior supports a model with many divergence-time parameters,
but we are unable to accurately approximate this posterior.
In Hypothesis~\ref{hypML}, the problem is more fundamental; given data from
taxa that diverged randomly through time, the exact posterior supports a model
with simultaneous divergences across taxa.
I.e., when accommodating prior uncertainty about divergence times with a
uniform distribution, the exact posterior from Bayes' rule leads us to the
wrong conclusion about evolutionary history.
Such posterior support for simultaneous divergence, even if ``correct'' from
the perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

@@

\begin{linenomath}
Although empirical Bayesian approaches can provide powerful methods for
parameter estimation, a theoretical justification for empirical Bayesian
approaches to model choice is questionable.
In Bayesian model choice, the primary goal is not to estimate parameters, but
to estimate the probabilities of candidate models.
In a simple example with two candidate models, $M_1$ and $M_2$, we can use
Bayes' rule to calculate the posterior probability of $M_1$ as
\begin{equation}
    p(M_1 \given \alignment{i}{}) = \frac{p(\alignment{i}{} \given
    M_1)p(M_1)}{p(\alignment{i}{} \given M_1)p(M_1) + p(\alignment{i}{} \given
    M_2)p(M_2)}.
    \label{eq:bayesmodelchoice}
\end{equation}
By comparing Equations \ref{eq:bayesrule} and \ref{eq:bayesmodelchoice}, we
see fundamental differences between Bayesian parameter estimation and
model choice.
\end{linenomath}

@@

However, \citet{Hickerson2013} conclude that the bias is caused by numerical-
approximation error due to insufficient computation
(Hypothesis~\ref{hypError}).
They argue the widest of the three priors on divergence times used by
\citet{Oaks2012} would infrequently produce random samples of parameter values
with many independent population divergence times as recent as the estimated
gene divergence times presented in \citet{Oaks2012}.
% They also also suggest that Oaks et al.'s \citeyear{Oaks2012} estimated gene
% divergence times should be used to set the upper bound of the uniform prior on
% divergence times for the \msb analysis of the same sequence data.
% Given the gene-tree estimates were intended to provide only a rough comparison
% of the gene divergence times across the 22 taxa \citep{Oaks2012}, and our
% findings presented above regarding the behavior of \msb when empirically
% informed uniform priors are employed, such an approach should not be used.
However, this sampling-probability argument is based on some questionable
assumptions.
Oaks et al.'s (\citeyear{Oaks2012}) gene-tree estimates were intended to
provide only a rough comparison of the gene divergence times across the 22 taxa
and assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations per
generation for all taxa.
Furthermore, because the branch-length units of the gene trees are in millions
of years whereas the divergence-time prior of \msb is in generations,
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
More importantly, even if we assume (1) the arbitrary strict clock is correct,
(2) gene divergence times were estimated without error, and (3) all 22 taxa
have one-year generation times, Hickerson et al.'s (\citeyear{Hickerson2013})
argument actually demonstrates that the models used by \citet{Oaks2012} with
narrower priors on divergence times are densely populated with samples with
large numbers of divergence parameters with values younger than the estimated
gene divergence estimates.
Thus, if \citet{Hickerson2013} are correct, analyses under these narrow priors
should be much less biased toward clustered models of divergence.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrowest prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug in
\msb, and after \citet{Oaks2012} corrected the bug, there are many cases where
the narrowest prior performs slightly worse (see panels D--J of Figures 3 and
S12).

@@

\section{Conclusions}
We demonstrate how the approximate-Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that often exclude the true
values of the divergence times.
Our results show that it is difficult to choose an uniformly distributed prior
on divergence times that is broad enough to confidently contain the true values
of parameters while being narrow enough to avoid strongly weighted and
misleading posterior support for models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection, because they can strongly influence the
posterior probabilities of the models we seek to compare.
No amount of computation can rescue our inference if our prior assumptions
place too much weight in unlikely regions of parameter space such that the
exact posterior supports the wrong model of evolutionary history.
